### 2021-05-23 chaps/thesis/chap1/sections/article.tex
🚚 rename thesis directory

Historically, particle in cell methods have been extensively used to treat kinetic problems. In this approach, the unknown is sampled by discrete particles which are advanced in time using an ODE solver, whereas the electric field is computed on a spatial grid. For some problems, these methods can tackle high dimensional kinetic problems with a relatively low computational cost. However, they also suffer from numerical noise which pollutes the accuracy in low density regions of phase space. Moreover, as the number of particles is increased the error only decreases as the inverse of the square of the number of particles. Thus, convergence is slow. For a review of particle methods we refer the reader to \cite{verboncoeur2005}.

@@

To overcome this CFL condition, semi-Lagrangian methods have been developed during the last decades. These methods realize a compromise between the Lagrangian (\ie~particle in cell) and Eulerian approaches by exploiting the characteristics equations to overcome the CFL condition, while still performing computations on a grid in both space and velocity \cite{cheng1976,sonnendrucker1999,filbet2003}. This approach is usually combined with splitting methods to avoid a costly multidimensional interpolation step. This allows for a separate treatment of the terms in the equation and the corresponding characteristic curves can then, at least in some situations, be computed analytically. For purely hyperbolic problems, the setting we consider here, it is also possible to construct high order splitting schemes \cite{casas2017}.

@@

In \cite{cep} an alternative approach based on exponential integrators was proposed. These schemes exploit the fact that in many applications where (gyro)kinetic models are used, the most stringent CFL condition is associated with the linear part of the model. This observation serves as the basis for the numerical methods we will consider in this work. Starting from the variation of constant formula, the linear part of the model will be solved exactly as part of an exponential integrator, whereas the nonlinear part, which is very often orders of magnitudes less stiff than the linear part, will be treated explicitly in time. In practice, the linear part can then be solved in phase space by using Fourier techniques or semi-Lagrangian schemes and the nonlinear part is approximated by standard finite difference/finite volume/discontinuous Galerkin techniques.

The numerical results presented in \cite{cep} were generally very favorable. The authors were able to take larger time steps compared to what has been reported for splitting methods in the literature and the computational cost was significantly reduced. In addition, since exponential integrators treat the nonlinear part explicitly, they can be adapted much more easily to different models. Despite these many favorable properties, the largest stable time step size was difficult to predict and varied significantly from method to method. Moreover, as we will see, many exponential integrators can behave rather erratically depending on the specific configuration of the simulation.

Thus, the main goal in this paper is to understand the stability of exponential integrators when applied to purely hyperbolic problems. While there is a large literature and well established theory for exponential integrators applied to parabolic problems (see \cite{ei} and references therein), we will see that for purely hyperbolic problems many surprises are encountered. Based on this analysis we will then propose to use a class of exponential methods, Lawson methods, that do not suffer from the described deficiency. Our analysis explains the efficient and robust behavior of Lawson integrators for this kind of problems. We will also present numerical results for both the Vlasov--Poisson equations and a drift-kinetic model that confirm the expected behavior and shows that using this approach significant performance improvements compared to the exponential integrators used in \cite{cep}, and by extension compared to splitting methods, can be attained.

@@

In this paper, we will consider two types of exponential methods. The idea of \textit{exponential integrators} is to use the variation of constants formula to rewrite equation \eqref{eq:expint-eq} in the following form
\[ u(t_n+\Delta t) = \exp(\Delta t A) u(t_n) + \int_0^{\Delta t} \exp((\Delta t -s)A) F(u(t_n+s)) \,\mathrm{d}s, \]
where we denote the time step size by $\Delta t >0$ and $t_n = n\Delta t$ with $n\in\mathbb{N}$. This expression is still exact; \ie no approximation has been made. Note, however, that this can not be used as a numerical method as evaluating the integral would require the knowledge of $u(t_n+s)$, which is not available. \textcolor{black}{The idea of an exponential integrator is to approximate the nonlinear part $F(u(t_n+s))$ in terms of the available data. In the simplest case we just evaluate it at the left endpoint. That is, we use $F(u(t_n+s))\approx F(u^n)$. Then we can integrate the term $\exp((\Delta t -s)A)$ exactly and obtain.}
\[ u(t_n+\Delta t) \approx u^{n+1} = \exp(\Delta t A) u^n + \Delta t \varphi_1(\Delta t A) F(u^n), \]
where $\varphi_1(z)=(\mathrm{e}^z-1)/z$ is an entire function. This is the first order exponential Euler method. In a similar way exponential Runge--Kutta methods can be constructed. We refer to the literature, in particular the review article \cite{ei}, for more details.

Another ansatz to remove the stiff linear term from equation \eqref{eq:expint-eq} is to introduce the change of variable
\[ v(t) := \exp(-t A) u(t). \]
Plugging this into equation \eqref{eq:expint-eq} yields
\begin{equation} \label{eq:lawson-eq} \dot{v}(t) = \exp(-t A) F(\exp(t A) v(t)). \end{equation}
Now we apply an explicit Runge--Kutta method to the transformed equation. In the simplest case, applying the explicit Euler scheme yields
\[ v(t_n+\Delta t) \approx v^{n+1} = v^{n} + \Delta t \exp(-t_n A) F(\exp(t_n A)v^{n}). \]
Reversing the change of variables yields
\[ u^{n+1} = \exp(\Delta t A) u^{n} + \Delta t \exp(\Delta t A) F(u^{n}). \]
This is the Lawson--Euler method, also a method of order one. Lawson methods are also commonly referred to as integrating factor methods. We immediately see that any explicit Runge--Kutta method applied to equation \eqref{eq:lawson-eq} uniquely determines a Lawson scheme. We call the chosen Runge--Kutta method the \textit{underlying Runge--Kutta method}. For more details we refer the reader to \cite{lawson1967,canuto1988,trefethen2000,minchev2005}.

@@

The efficiency of exponential methods crucially depends on a good method to evaluate the application of the required matrix functions to a vector. A range of methods has been developed to accomplish this. For example, Krylov methods or interpolation at Leja points can be used for a wide range of problems; see, for example, \cite{higham2008,hochbruck1997,al2011,caliari2014, ckor15}. 
 However, often the most efficient approach is to exploit particular knowledge about the differential equation under consideration. For example, in the hyperbolic case $A$ might be a linear advection operator. In this case the application of $\exp(\Delta t A)$ can be computed by using Fourier techniques or semi-Lagrangian schemes. Much research effort has been dedicated towards improving spectral and semi-Lagrangian schemes for kinetic problems \cite{crouseilles2011,einkemmer2014,einkemmer2016,filbet2003,grandgirard,klimas1994,Morrison2017,rossmanith2011,sonnendrucker1999,cheng1976,sircombe2009valis,crouseilles2015hamiltonian,crouseilles2016asymptotic,einkemmer2019comparison,einkemmer2014convergence,einkemmer2014dG} and obtaining good performance on state of the art HPC systems \cite{rozar2013,einkemmer2015,bigot2013,latu2007gyrokinetic,mehrenberger2013vlasov,einkemmer2016mixed,crouseilles2009parallel,einkemmer2019gpu}.

Before proceeding, let us note that for parabolic problems, \ie~where $A$ is an elliptic operator, a mature theory for exponential integrators is available. We again refer the reader to the review article \cite{ei}. In this setting there are relatively few surprises with respect to stability and even rigorous convergence results are available. In addition, exponential integrators have been considered for problems that include both hyperbolic and parabolic terms (see, for example \cite{martinez2009,tambue2010,einkemmer2016expintmhd,einkemmer2013expintgpu}). An interesting point to make is that in this community Lawson methods have all but lost their appeal. In fact, there are many reasons why exponential integrators are to be preferred. For example, if a Krylov method is used to compute the matrix functions, the $\varphi_1$ function usually converges faster than the exponential. In addition, exponential integrators that retain their full order for non-homogeneous boundary conditions have been constructed~\cite{hochbruck2005}. It has been shown that this property can not be achieved for Lawson methods \cite{hochbruck2017lawson}. However, the situation for purely hyperbolic problems is markedly different. Most of the theoretical results that have been obtained in an abstract framework do not apply and there is relatively little literature available. We will see  that the stability for exponential integrators in the fully hyperbolic setting is full of surprises. Moreover, since for kinetic problems we usually have efficient methods to compute the matrix exponential and complicated boundary conditions are rather rare, Lawson methods are an attractive choice due to their improved stability, as we will see.

@@

This makes the problem rather easy as the stability constraint for explicit Runge--Kutta methods has been extensively studied in the literature. In our present application we are primarily interested in obtaining numerical methods that maximize the part of the imaginary axis that is included in the domain of stability. It is well known that an $s$ stage method can include at most $i [-(s-1),s-1]$. This is, for example, stated as an exercise in \cite[Chapter. IV.2, exercise 3]{sode}. Thus, unfortunately, there is no analog to Runge--Kutta--Chebyshev methods for hyperbolic problems.

@@

We now apply commonly used exponential integrators to the test equation \eqref{ode_linear}. In this work we will consider the following methods: ExpRK22 (a classic two stage second order method), the method of Cox--Matthews \cite{cox}, the method of Hochbruck--Ostermann  \cite{hochbruck2005}, and the method of Krogstad \cite{krogstad2005}. We refer to \cite{ei} for more details and to Appendix \ref{butcher} for the Butcher tableaus of these methods.

@@

A common technique to discretize hyperbolic partial differential equations is to use the so-called weighted essentially non-oscillatory schemes (WENO) schemes. These are nonlinear schemes that limit oscillations in regions where sharp gradients occur, but still yield high order accuracy in smooth regions of the phase space. In the linear case WENO schemes reduce to upwind discretizations. Here, we will consider the LW5 scheme (the linearized version of the WENO5 scheme as considered in \cite{baldauf, lunet, motamed, wang}) that is given by (from now on we assume w.l.o.g.~that $b>0$)
\begin{align*}
(\partial_v \hat{f}_k)(v_j) &\approx \frac{1}{\Delta v}\Big(-\frac{1}{30} \hat{f}_{k,j-3} +\frac{1}{4} \hat{f}_{k,j-2} -\hat{f}_{k,j-1} + \frac{1}{3} \hat{f}_{k,j} +\frac{1}{2} \hat{f}_{k,j+1} - \frac{1}{20} \hat{f}_{k,j+2}\Big). 
\end{align*}
We now perform the same analysis as for the centered scheme (see \cite{baldauf, fov}). This yields
\begin{align}
  {\color{black} \left( \frac{2\pi }{2 v_{\max}}  i m \bar{f}_{k, m} \approx \right)}
   %(i m \bar{f}_{k, m} \approx ) \;\; 
   \mu_m \bar{f}_{k, m} &:= \frac{\bar{f}_{k,m}}{\Delta v}\Big(-\frac{1}{30} e^{-\frac{3i m\pi \Delta v}{v_{\max}}} +\frac{1}{4} e^{-\frac{2i m \pi \Delta v}{v_{\max}}} -e^{-\frac{i m \pi \Delta v}{v_{\max}}}  \nonumber\\
&\hspace{1cm}+ \frac{1}{3}  +\frac{1}{2} e^{\frac{i m \pi \Delta v}{v_{\max}}} - \frac{1}{20} e^{\frac{2i m \pi \Delta v}{v_{\max}}}\Big). 
\label{lw5symbol}
\end{align}
We then obtain the equation 
\begin{equation}
	\label{discrete_linear_transport_weno}
	\partial_t \bar{f}_{k,m} + i dk\bar{f}_{k,m} +b \mu_m \bar{f}_{k,m}= 0.
\end{equation}
Once again this is precisely the form of equation \eqref{ode_linear}, where $a=dk\in\mathbb{R}$ and $\lambda=b\mu_m \in\mathbb{C}$. The main difference to the centered difference scheme is that $\lambda$ is not necessarily on the imaginary axis. In fact, the eigenvalues aquire a negative real part which stabilizes the scheme and avoids spurious oscillations, but also adds unphysical dissipation to the numerical method.

@@

For the exponential integrators the domain of stability is very sensitive to the value of $(a\Delta t)$. To get an idea of what we can expect, we consider the quantity $y_{\max} = \min_{(a\Delta t)\in\mathbb{R}} \; y^{exp}_{\max}(a \Delta t)$. As before, $y^{exp}_{\max}(a\Delta t)$ is the largest value such that $i(-y^{exp}_{\max}(a\Delta t), y^{exp}_{\max}(a\Delta t))\subset {\cal D}$, where ${\cal D}$ is the domain of stability for the chosen exponential time integrator for a given $(a\Delta t)$. 
Even for relatively simple numerical methods it is not possible to compute this quantity analytically. Thus, we resort to numerical approximations. Unfortunately, it turns out that for most exponential integrators this value is zero. This can be appreciated by considering Figure \ref{fig:expRK_sd} once more. Clearly, there are values of $(a \Delta t)$ such that no relevant part of the imaginary axis (or only half the imaginary axis) is part of the domain of stability.
Thus, most exponential integrators are unstable in the von Neumann sense. However, this is not what we observe in practice. In fact, already the results presented in \cite{cep} indicate that we can successfully run numerical simulations using, for example, the Cox--Matthews scheme. There are two major points to consider here
\begin{itemize}
    \item The $y_{\max}$ obtained is a worst case estimate. In fact, we know that for $\Delta t \to 0$ we regain the stability of the underlying Runge--Kutta method. Thus, for small $(a \Delta t)$ the methods is expected to work well.
    \item As is usually done we have mandated that $\vert \phi(z) \vert \leq 1$. However, strictly speaking this is not necessary for practical simulation. If we assume that $\vert \phi(z) \vert \leq 1+\varepsilon$ and we take $n$ steps the amplification of the error is given by $(1+\varepsilon)^n$. In the limit $n \to +\infty$ this quantity diverges. However, since we usually do not take infinitely small time steps we still can hope to obtain a relatively accurate approximation, especially if $\varepsilon$ is small. \textcolor{black}{In particular, if $\varepsilon = C \Delta t = C t_{final}/n$ (where $t_{final}$ denotes the final time and $n$ the number of iterations) we have $(1+\varepsilon)^n = (1+\tfrac{C t_{final}}{n} )^n \leq \exp(C t_{final})$ and thus the scheme is stable, while the error constant is increased by $\exp(C t_{final})$.}
\end{itemize}

@@

In the case of a WENO5 approximation of the velocity derivative, we can not 
easily find a Fourier multiplier because of its nonlinearity. 
Recent studies about stability of WENO5 \cite{wang, motamed, lunet} considered 
the linearized version of WENO schemes by freezing the nonlinear weights, 
so that WENO5 reduces to a high order (linear) upwind scheme called LW5. For this LW5 scheme we can
compute the eigenvalues, see equation \eqref{lw5symbol}, 
and different time stepping schemes can be studied to determine the stability limit. 
We consider only Lawson methods here since we found 
that the exponential schemes we considered (\ie{}~ExpRK22, Krogstad, Cox--Matthews, Hochbruck--Ostermann) 
lead to unstable results when they are combined with LW5 (even in the weak sense considered in the previous section).

@@

In Figure \ref{total_energy}, we plot the time evolution of $({\cal H}^n -{\cal H}(0))/{\cal H}(0)$, where ${\cal H}^n\approx {\cal H}(n\Delta t)$ and ${\cal H}(t)$ is the total energy defined by 
$$
{\cal H}(t) = \frac{1}{2}\int\int |v|^2 f(t, x, v) \,dxdv + \frac{1}{2}\int |E|^2(t, x) \,dx. 
$$
This quantity is known to be preserved with time {\color{black} at the continuous level. It is thus a useful metric to evaluate and compare the different numerical 
methods. At this stage, all the used numerical methods are stable and we now look at their accuracy with respect to conservation of energy. }
% and thus allows us to look at the accuracy of the different methods considered in this paper.
We observe that Lawson/centered schemes (referred as 'CD2' in the 
legend) preserve this quantity well. It is well known (see for example \cite{cf}) that centered schemes are better at preserving the total energy compared to upwind schemes. The reason is that upwind schemes introduce numerical diffusion. The exponential integrators that are considered show all very similar behavior with respect to energy conservation. They seem to include less drift than the Lawson methods, but for the time scales considered here their error is larger.

@@

Our goal in this section is to find a numerical approximation of $f=f(t,r,\theta,z,v)$ satisfying the following $4D$ slab drift-kinetic equation (see \cite{grandgirard})
\begin{equation}
\label{dk}
\partial_tf-\frac{\partial_\theta \phi}{r}\partial_rf+\frac{\partial_r \phi}{r}\partial_\theta f
+v\partial_zf-\partial_z\phi\partial_{v}f=0,
\end{equation}
for $(r,\theta,z,v)\in \Omega\times[0,L]\times \mathbb{R}$, $\Omega=[r_{\rm min},r_{\rm max}]\times [0, 2\pi]$.
The self-consistent potential $\phi=\phi(r,\theta,z)$ is determined by solving the quasi neutrality equation
\begin{align}
    -\left[\partial_r^2\phi+\left(\frac{1}{r}+\frac{\partial_r n_0(r)}{n_0(r)}\right)\partial_r\phi \right.&+\left.\frac{1}{r^2}\partial_\theta^2\phi\right]+\frac{1}{T_e(r)}(\phi-\langle\phi\rangle)\nonumber\\
\label{qn}
&\hspace{-3cm}=\frac{1}{n_0(r)}\int_{\mathbb{R}} fdv-1,
\end{align}
where $\langle\phi\rangle = \frac{1}{L}\int_0^L \phi(r,\theta,z)\,dz$ and the functions $n_0$ and $T_e$ depend only on $r$ and are given analytically.

>>>>

Historically, particle in cell methods have been extensively used to treat kinetic problems. In this approach, the unknown is sampled by discrete particles which are advanced in time using an ODE solver, whereas the electric field is computed on a spatial grid. For some problems, these methods can tackle high dimensional kinetic problems with a relatively low computational cost. However, they also suffer from numerical noise which pollutes the accuracy in low density regions of phase space. Moreover, as the number of particles is increased the error only decreases as the inverse of the square of the number of particles. Thus, convergence is slow. For a review of particle methods we refer the reader to \cite{Verboncoeur:2005}.

@@

To overcome this CFL condition, semi-Lagrangian methods have been developed during the last decades. These methods realize a compromise between the Lagrangian (\ie~particle in cell) and Eulerian approaches by exploiting the characteristics equations to overcome the CFL condition, while still performing computations on a grid in both space and velocity \cite{Cheng:1976,Sonnendrucker:1999,Filbet:2003}. This approach is usually combined with splitting methods to avoid a costly multidimensional interpolation step. This allows for a separate treatment of the terms in the equation and the corresponding characteristic curves can then, at least in some situations, be computed analytically. For purely hyperbolic problems, the setting we consider here, it is also possible to construct high order splitting schemes \cite{Casas:2017}.

@@

In \cite{Crouseilles:2018} an alternative approach based on exponential integrators was proposed. These schemes exploit the fact that in many applications where (gyro)kinetic models are used, the most stringent CFL condition is associated with the linear part of the model. This observation serves as the basis for the numerical methods we will consider in this work. Starting from the variation of constant formula, the linear part of the model will be solved exactly as part of an exponential integrator, whereas the nonlinear part, which is very often orders of magnitudes less stiff than the linear part, will be treated explicitly in time. In practice, the linear part can then be solved in phase space by using Fourier techniques or semi-Lagrangian schemes and the nonlinear part is approximated by standard finite difference/finite volume/discontinuous Galerkin techniques.

The numerical results presented in \cite{Crouseilles:2018} were generally very favorable. The authors were able to take larger time steps compared to what has been reported for splitting methods in the literature and the computational cost was significantly reduced. In addition, since exponential integrators treat the nonlinear part explicitly, they can be adapted much more easily to different models. Despite these many favorable properties, the largest stable time step size was difficult to predict and varied significantly from method to method. Moreover, as we will see, many exponential integrators can behave rather erratically depending on the specific configuration of the simulation.

Thus, the main goal in this paper is to understand the stability of exponential integrators when applied to purely hyperbolic problems. While there is a large literature and well established theory for exponential integrators applied to parabolic problems (see \cite{Hochbruck:2010} and references therein), we will see that for purely hyperbolic problems many surprises are encountered. Based on this analysis we will then propose to use a class of exponential methods, Lawson methods, that do not suffer from the described deficiency. Our analysis explains the efficient and robust behavior of Lawson integrators for this kind of problems. We will also present numerical results for both the Vlasov--Poisson equations and a drift-kinetic model that confirm the expected behavior and shows that using this approach significant performance improvements compared to the exponential integrators used in \cite{Crouseilles:2018}, and by extension compared to splitting methods, can be attained.

@@

In this paper, we will consider two types of exponential methods. The idea of \textit{exponential integrators} is to use the variation of constants formula to rewrite equation \eqref{eq:expint-eq} in the following form
\[ u(t_n+\Delta t) = \exp(\Delta t A) u(t_n) + \int_0^{\Delta t} \exp((\Delta t -s)A) F(u(t_n+s)) \,\mathrm{d}s, \]
where we denote the time step size by $\Delta t >0$ and $t_n = n\Delta t$ with $n\in\mathbb{N}$. This expression is still exact; \ie no approximation has been made. Note, however, that this can not be used as a numerical method as evaluating the integral would require the knowledge of $u(t_n+s)$, which is not available. \textcolor{black}{The idea of an exponential integrator is to approximate the nonlinear part $F(u(t_n+s))$ in terms of the available data. In the simplest case we just evaluate it at the left endpoint. That is, we use $F(u(t_n+s))\approx F(u^n)$. Then we can integrate the term $\exp((\Delta t -s)A)$ exactly and obtain.}
\[ u(t_n+\Delta t) \approx u^{n+1} = \exp(\Delta t A) u^n + \Delta t \varphi_1(\Delta t A) F(u^n), \]
where $\varphi_1(z)=(\mathrm{e}^z-1)/z$ is an entire function. This is the first order exponential Euler method. In a similar way exponential Runge--Kutta methods can be constructed. We refer to the literature, in particular the review article \cite{Hochbruck:2010}, for more details.

Another ansatz to remove the stiff linear term from equation \eqref{eq:expint-eq} is to introduce the change of variable
\[ v(t) := \exp(-t A) u(t). \]
Plugging this into equation \eqref{eq:expint-eq} yields
\begin{equation} \label{eq:lawson-eq} \dot{v}(t) = \exp(-t A) F(\exp(t A) v(t)). \end{equation}
Now we apply an explicit Runge--Kutta method to the transformed equation. In the simplest case, applying the explicit Euler scheme yields
\[ v(t_n+\Delta t) \approx v^{n+1} = v^{n} + \Delta t \exp(-t_n A) F(\exp(t_n A)v^{n}). \]
Reversing the change of variables yields
\[ u^{n+1} = \exp(\Delta t A) u^{n} + \Delta t \exp(\Delta t A) F(u^{n}). \]
This is the Lawson--Euler method, also a method of order one. Lawson methods are also commonly referred to as integrating factor methods. We immediately see that any explicit Runge--Kutta method applied to equation \eqref{eq:lawson-eq} uniquely determines a Lawson scheme. We call the chosen Runge--Kutta method the \textit{underlying Runge--Kutta method}. For more details we refer the reader to \cite{Lawson:1967a,Canuto:1988,Trefethen:2000,Minchev:2005}.

@@

The efficiency of exponential methods crucially depends on a good method to evaluate the application of the required matrix functions to a vector. A range of methods has been developed to accomplish this. For example, Krylov methods or interpolation at Leja points can be used for a wide range of problems; see, for example, \cite{Higham:2008,Hochbruck:1997,Al-Mohy:2011,Caliari:2014, Caliari:2016}. 
 However, often the most efficient approach is to exploit particular knowledge about the differential equation under consideration. For example, in the hyperbolic case $A$ might be a linear advection operator. In this case the application of $\exp(\Delta t A)$ can be computed by using Fourier techniques or semi-Lagrangian schemes. Much research effort has been dedicated towards improving spectral and semi-Lagrangian schemes for kinetic problems \cite{Crouseilles:2011,Einkemmer:2014,Einkemmer:2017a,Filbet:2003,Grandgirard:2006,Klimas:1994,Morrison:2017,Rossmanith:2011,Sonnendrucker:1999,Cheng:1976,Sircombe:2009,Crouseilles:2015,Crouseilles:2016,Einkemmer:2019,Einkemmer:2014b,Einkemmer:2014a} and obtaining good performance on state of the art HPC systems \cite{Rozar:2014,Einkemmer:2015,Bigot:2012,Latu:2007,Mehrenberger:2013,Einkemmer:2016,Crouseilles:2009,Einkemmer:2020}.

Before proceeding, let us note that for parabolic problems, \ie~where $A$ is an elliptic operator, a mature theory for exponential integrators is available. We again refer the reader to the review article \cite{Hochbruck:2010}. In this setting there are relatively few surprises with respect to stability and even rigorous convergence results are available. In addition, exponential integrators have been considered for problems that include both hyperbolic and parabolic terms (see, for example \cite{Martinez:2009,Tambue:2010,Einkemmer:2017,Einkemmer:2013}). An interesting point to make is that in this community Lawson methods have all but lost their appeal. In fact, there are many reasons why exponential integrators are to be preferred. For example, if a Krylov method is used to compute the matrix functions, the $\varphi_1$ function usually converges faster than the exponential. In addition, exponential integrators that retain their full order for non-homogeneous boundary conditions have been constructed~\cite{Hochbruck:2005}. It has been shown that this property can not be achieved for Lawson methods \cite{Hochbruck:2020}. However, the situation for purely hyperbolic problems is markedly different. Most of the theoretical results that have been obtained in an abstract framework do not apply and there is relatively little literature available. We will see  that the stability for exponential integrators in the fully hyperbolic setting is full of surprises. Moreover, since for kinetic problems we usually have efficient methods to compute the matrix exponential and complicated boundary conditions are rather rare, Lawson methods are an attractive choice due to their improved stability, as we will see.

@@

This makes the problem rather easy as the stability constraint for explicit Runge--Kutta methods has been extensively studied in the literature. In our present application we are primarily interested in obtaining numerical methods that maximize the part of the imaginary axis that is included in the domain of stability. It is well known that an $s$ stage method can include at most $i [-(s-1),s-1]$. This is, for example, stated as an exercise in \cite[Chapter. IV.2, exercise 3]{Hairer:1996}. Thus, unfortunately, there is no analog to Runge--Kutta--Chebyshev methods for hyperbolic problems.

@@

We now apply commonly used exponential integrators to the test equation \eqref{ode_linear}. In this work we will consider the following methods: ExpRK22 (a classic two stage second order method), the method of Cox--Matthews \cite{Cox:2002}, the method of Hochbruck--Ostermann  \cite{Hochbruck:2005}, and the method of Krogstad \cite{Krogstad:2005}. We refer to \cite{Hochbruck:2010} for more details and to Appendix \ref{butcher} for the Butcher tableaus of these methods.

@@

A common technique to discretize hyperbolic partial differential equations is to use the so-called weighted essentially non-oscillatory schemes (WENO) schemes. These are nonlinear schemes that limit oscillations in regions where sharp gradients occur, but still yield high order accuracy in smooth regions of the phase space. In the linear case WENO schemes reduce to upwind discretizations. Here, we will consider the LW5 scheme (the linearized version of the WENO5 scheme as considered in \cite{Baldauf:2008, Lunet:2017, Motamed:2010, Wang:2007}) that is given by (from now on we assume w.l.o.g.~that $b>0$)
\begin{align*}
(\partial_v \hat{f}_k)(v_j) &\approx \frac{1}{\Delta v}\Big(-\frac{1}{30} \hat{f}_{k,j-3} +\frac{1}{4} \hat{f}_{k,j-2} -\hat{f}_{k,j-1} + \frac{1}{3} \hat{f}_{k,j} +\frac{1}{2} \hat{f}_{k,j+1} - \frac{1}{20} \hat{f}_{k,j+2}\Big). 
\end{align*}
We now perform the same analysis as for the centered scheme (see \cite{Baldauf:2008, Crouseilles:2012}). This yields
\begin{align}
  {\color{black} \left( \frac{2\pi }{2 v_{\max}}  i m \bar{f}_{k, m} \approx \right)}
   %(i m \bar{f}_{k, m} \approx ) \;\; 
   \mu_m \bar{f}_{k, m} &:= \frac{\bar{f}_{k,m}}{\Delta v}\Big(-\frac{1}{30} e^{-\frac{3i m\pi \Delta v}{v_{\max}}} +\frac{1}{4} e^{-\frac{2i m \pi \Delta v}{v_{\max}}} -e^{-\frac{i m \pi \Delta v}{v_{\max}}}  \nonumber\\
&\hspace{1cm}+ \frac{1}{3}  +\frac{1}{2} e^{\frac{i m \pi \Delta v}{v_{\max}}} - \frac{1}{20} e^{\frac{2i m \pi \Delta v}{v_{\max}}}\Big). 
\label{lw5symbol}
\end{align}
We then obtain the equation 
\begin{equation}
	\label{discrete_linear_transport_weno}
	\partial_t \bar{f}_{k,m} + i dk\bar{f}_{k,m} +b \mu_m \bar{f}_{k,m}= 0.
\end{equation}
Once again this is precisely the form of equation \eqref{ode_linear}, where $a=dk\in\mathbb{R}$ and $\lambda=b\mu_m \in\mathbb{C}$. The main difference to the centered difference scheme is that $\lambda$ is not necessarily on the imaginary axis. In fact, the eigenvalues aquire a negative real part which stabilizes the scheme and avoids spurious oscillations, but also adds unphysical dissipation to the numerical method.

@@

For the exponential integrators the domain of stability is very sensitive to the value of $(a\Delta t)$. To get an idea of what we can expect, we consider the quantity $y_{\max} = \min_{(a\Delta t)\in\mathbb{R}} \; y^{exp}_{\max}(a \Delta t)$. As before, $y^{exp}_{\max}(a\Delta t)$ is the largest value such that $i(-y^{exp}_{\max}(a\Delta t), y^{exp}_{\max}(a\Delta t))\subset {\cal D}$, where ${\cal D}$ is the domain of stability for the chosen exponential time integrator for a given $(a\Delta t)$. 
Even for relatively simple numerical methods it is not possible to compute this quantity analytically. Thus, we resort to numerical approximations. Unfortunately, it turns out that for most exponential integrators this value is zero. This can be appreciated by considering Figure \ref{fig:expRK_sd} once more. Clearly, there are values of $(a \Delta t)$ such that no relevant part of the imaginary axis (or only half the imaginary axis) is part of the domain of stability.
Thus, most exponential integrators are unstable in the von Neumann sense. However, this is not what we observe in practice. In fact, already the results presented in \cite{Crouseilles:2018} indicate that we can successfully run numerical simulations using, for example, the Cox--Matthews scheme. There are two major points to consider here
\begin{itemize}
    \item The $y_{\max}$ obtained is a worst case estimate. In fact, we know that for $\Delta t \to 0$ we regain the stability of the underlying Runge--Kutta method. Thus, for small $(a \Delta t)$ the methods is expected to work well.
    \item As is usually done we have mandated that $\vert \phi(z) \vert \leq 1$. However, strictly speaking this is not necessary for practical simulation. If we assume that $\vert \phi(z) \vert \leq 1+\varepsilon$ and we take $n$ steps the amplification of the error is given by $(1+\varepsilon)^n$. In the limit $n \to +\infty$ this quantity diverges. However, since we usually do not take infinitely small time steps we still can hope to obtain a relatively accurate approximation, especially if $\varepsilon$ is small. \textcolor{black}{In particular, if $\varepsilon = C \Delta t = C t_{final}/n$ (where $t_{final}$ denotes the final time and $n$ the number of iterations) we have $(1+\varepsilon)^n = (1+\tfrac{C t_{final}}{n} )^n \leq \exp(C t_{final})$ and thus the scheme is stable, while the error constant is increased by $\exp(C t_{final})$.}
\end{itemize}

@@

In the case of a WENO5 approximation of the velocity derivative, we can not 
easily find a Fourier multiplier because of its nonlinearity. 
Recent studies about stability of WENO5 \cite{Wang:2007, Motamed:2010, Lunet:2017} considered 
the linearized version of WENO schemes by freezing the nonlinear weights, 
so that WENO5 reduces to a high order (linear) upwind scheme called LW5. For this LW5 scheme we can
compute the eigenvalues, see equation \eqref{lw5symbol}, 
and different time stepping schemes can be studied to determine the stability limit. 
We consider only Lawson methods here since we found 
that the exponential schemes we considered (\ie{}~ExpRK22, Krogstad, Cox--Matthews, Hochbruck--Ostermann) 
lead to unstable results when they are combined with LW5 (even in the weak sense considered in the previous section).

@@

In Figure \ref{total_energy}, we plot the time evolution of $({\cal H}^n -{\cal H}(0))/{\cal H}(0)$, where ${\cal H}^n\approx {\cal H}(n\Delta t)$ and ${\cal H}(t)$ is the total energy defined by 
$$
{\cal H}(t) = \frac{1}{2}\int\int |v|^2 f(t, x, v) \,dxdv + \frac{1}{2}\int |E|^2(t, x) \,dx. 
$$
This quantity is known to be preserved with time {\color{black} at the continuous level. It is thus a useful metric to evaluate and compare the different numerical 
methods. At this stage, all the used numerical methods are stable and we now look at their accuracy with respect to conservation of energy. }
% and thus allows us to look at the accuracy of the different methods considered in this paper.
We observe that Lawson/centered schemes (referred as 'CD2' in the 
legend) preserve this quantity well. It is well known (see for example \cite{Crouseilles:2004}) that centered schemes are better at preserving the total energy compared to upwind schemes. The reason is that upwind schemes introduce numerical diffusion. The exponential integrators that are considered show all very similar behavior with respect to energy conservation. They seem to include less drift than the Lawson methods, but for the time scales considered here their error is larger.

@@

Our goal in this section is to find a numerical approximation of $f=f(t,r,\theta,z,v)$ satisfying the following $4D$ slab drift-kinetic equation (see \cite{Grandgirard:2006})
\begin{equation}
\label{dk}
\partial_tf-\frac{\partial_\theta \phi}{r}\partial_rf+\frac{\partial_r \phi}{r}\partial_\theta f
+v\partial_zf-\partial_z\phi\partial_{v}f=0,
\end{equation}
for $(r,\theta,z,v)\in \Omega\times[0,L]\times \mathbb{R}$, $\Omega=[r_{\rm min},r_{\rm max}]\times [0, 2\pi]$.
The self-consistent potential $\phi=\phi(r,\theta,z)$ is determined by solving the quasi neutrality equation
\begin{align}
    -\left[\partial_r^2\phi+\left(\frac{1}{r}+\frac{\partial_r n_0(r)}{n_0(r)}\right)\partial_r\phi \right.&+\left.\frac{1}{r^2}\partial_\theta^2\phi\right]+\frac{1}{T_e(r)}(\phi-\langle\phi\rangle)\nonumber\\
\label{qn}
&\hspace{-3cm}=\frac{1}{n_0(r)}\int_{\mathbb{R}} fdv-1,
\end{align}
where $\langle\phi\rangle = \frac{1}{L}\int_0^L \phi(r,\theta,z)\,dz$ and the functions $n_0$ and $T_e$ depend only on $r$ and are given analytically.

In many situations the $v \partial_z f$ term yields the most restrictive CFL condition. In this setting exponential methods can be very successful as they remove the most stringent CFL condition, while still treating the remaining terms explicitly (which computationally is relatively cheap). The $\varphi$ functions can be computed in Fourier space (as has been discussed in some detail for the Vlasov--Poisson system in the previous section) or using a semi-Lagrangian approach. Exponential integrators for the drift-kinetic model have been proposed in \cite{Crouseilles:2018}. They compare favorably to splitting schemes and have the advantage that they can be more easily adapted to different models. In \cite{Crouseilles:2018} only a second order exponential integrator and the fourth order Cox--Matthews scheme have been considered. Due to the investigations in the present paper we now understand that this is not an ideal choice. Thus, the purpose of this section is to demonstrate that Lawson methods can be more efficient and to further corroborate the results obtained in the previous sections. The difference in stability for Lawson schemes and exponential integrators will be very evident in the numerical simulations that are presented.
